---
title: "Work document: Alzheimer demo data"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Alzheimer dataset descriptives. Target of this demo project is to identify factors/variables associate with Alzheimer disease in KUH data lake demo dataset. Also interest is to find out if there is predictive power in these factors/variables to identify upcoming Alzheimer disease diagnose. Project is available in Gitlab <https://github.com/pohjois-savon-tietoallas/demo-alzheimer-shiny>.


# Datasets
 
```{r}
library(tidyverse)
library(dplyr) # required for data wrangling mutate (part of library(tidyverse))
library(Hmisc) # required for Hmisc::describe function (Imporant to load this first. Otherwise the dplyr::summarize will be overridden.)
library(tidyverse) # tools for data wrangling
library(lubridate) # required for Date manipulation
library(wordcloud)
library(RColorBrewer)

load(file="/dbfs/mnt/demodata/norpen.RData", verbose=T)
```

## Persons

 - Includes demographics of "persons" who have obtained a special reimbursement for medications related to AD during 2005-2012
 
```{r}
str(persons) # 13 431
summary(persons)
```
 
 
## Special Reimbursements

 - data from the Special Reimbursement Register
 - Entitlements to higher medication reimbursement due to chronic diseases with start and end dates and classification numbers
 
```{r}
str(srr)
summary(srr)
```

Diagnoses 


```{r}
d1 <- srr %>% group_by(diag) %>% summarise(n=n()) %>% filter(diag != '')
wordcloud(words = d1$diag, freq = d1$n, colors = c("#999999", "#E69F00", "#56B4E9","#56B4E9"))
```

```{r}
d1 <- srr %>% group_by(sk1) %>% summarise(n=n())
wordcloud(words = d1$sk1, freq = d1$n, colors = c("#999999", "#E69F00", "#56B4E9","#56B4E9"))
```


 
## drugs
 
 - Reimbursed medications from the Finnish Prescription Register
 - ATC code, dispensing date, strength, number of tablets dispensed

```{r}
str(drugs) # 402 180
summary(drugs)
```

 
## care_reg

 - data from the Finnish Care Register
 - primary and secondary discharge diagnoses, medical procedures, admission and discharge dates
 - in ICD-8 (admissions in 1972-1986), ICD-9 (1987-1995) and ICD-10 (1996->) formats

```{r}
str(care_reg)
summary(care_reg)
```

Variables:
 
 - idnum = identification number
 - adm_date = date of admission
 - disch_date = date of discharge
 - pdgo = primary diagnosis (ICD-10)
 - pdge = primary diagnosis (ICD-10)
 - sdg1o = secondary diagnosis (ICD-10)
 - sdg1e = secondary diagnosis (ICD-10)
 - sdg2o = secondary diagnosis (ICD-10)
 - sdg2e = secondary diagnosis (ICD-10)
 - ptmpk1 = surgical procedure (NOMESCO)
 - ptmpk2 = surgical procedure (NOMESCO)
 - ptmpk3 = surgical procedure (NOMESCO)
 - mtmp1k1 = surgical procedure (NOMESCO)
 - mtmp2k1 = surgical procedure (NOMESCO)
 - pdg = primary diagnosis (ICD-9)
 - sdg1 = 1st secondary diagnosis (ICD-9)
 - sdg2 = 2nd secondary diagnosis (ICD-9)
 - dg1 = diagnosis (ICD-8)
 - dg2 = diagnosis (ICD-8)
 - dg3 = diagnosis (ICD-8)
 - dg4 = diagnosis (ICD-8)


## inst_reg

 - long-term institutionalizations from the SII's register
 - start and end dates

```{r}
str(inst_reg)
summary(inst_reg)
```


```{r}
# hoitojaksot jotka olleet ennen diagnoosia (alku ja lopppu ennen diagnoosi), kestot
inst_reg %>% 
  left_join(persons) %>% 
  mutate(pv1 = erko307_pv - lhjalkpv,
         pv2 = erko307_pv - lhjlakpv,
         inst_reg_days = lhjlakpv - lhjalkpv) %>% 
  filter(pv1 > 0 & pv2 > 0) %>% 
  select(idnum, pv1, inst_reg_days) -> dplot

ggplot(dplot) +
  geom_point(aes(x=pv1, y=inst_reg_days)) 
```


## mort

 - data from the Causes of Death register
 - date of death
 
```{r}
str(mort)
summary(mort)
```





# Explorative data analysis

## Persons

```{r}
# Persons data join ----
persons <- left_join(persons, mort)
# classify vars
persons$sp <- factor(persons$sp, labels=c("Male", "Female"))
# First, start by calculating age at the time of AD diagnosis;
persons <- mutate(persons,
                  alz_age=as.numeric((erko307_pv-syntpv)/365.25),
                  alz_age_yrs=as.numeric(round(alz_age)),
                  erko_yr=year(erko307_pv)  #  or =format(erko307_pv,"%Y")  # year of AD diagnosis
)
# head(persons)
summary(persons)
```

```{r}
# age distruribution
d1 <- persons %>% group_by(format(persons$syntpv, "%Y"), sp) %>% summarise(n=n())
d1$year <- d1$`format(persons$syntpv, "%Y")`
ggplot(d1) +
  geom_bar(stat = "identity", aes(x=year, y=n, fill=sp)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Birth year distribution") +
  scale_fill_manual(values=c("dodgerblue3", "orchid2"))
```

```{r}
# gender
d1 <- persons %>% group_by(sp) %>% summarise(n=n())
ggplot(d1) +
  geom_bar(aes(x=sp, y=n, fill=sp), stat = "identity") +
  geom_text(aes(x=sp, y=n, label = n)) +
  labs(title = "Gender distribution") + 
  scale_fill_manual(values=c("dodgerblue3", "orchid2"))
```

```{r}
# mortality
d1 <- persons %>% group_by(format(persons$kuolpv, "%Y"), sp) %>% summarise(n=n())
d1$year <- d1$`format(persons$kuolpv, "%Y")`
ggplot(d1[!is.na(d1$year),]) +
  geom_bar(stat = "identity", aes(x=year, y=n, fill=sp)) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mortality, year") +
  scale_fill_manual(values=c("dodgerblue3", "orchid2"))
```


```{r}
# mortality after diagnose
d1 <- persons %>% 
  mutate(died_after_diag = kuolpv - erko307_pv) %>% 
  filter(!is.na(died_after_diag)) %>% 
  group_by(died_after_diag, sp) %>% 
  summarise(n=n()) %>% 
  mutate(
    died_after_diag_cat = cut(as.integer(died_after_diag),
                              breaks=c(-Inf, 180, 365, 730, 1095, Inf),
                              labels = c("0-180 days", "181-365 days", "366-730 days", "730-1095 days", "1095+")
                              )
  )
ggplot(d1) +
  geom_bar(stat = "identity", aes(x=died_after_diag_cat, y=n, fill=sp)) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mortality, year") +
  scale_fill_manual(values=c("dodgerblue3", "orchid2"))
```


```{r}
# What age is alzheimer disease diagnosed
d1 <- persons %>% group_by(alz_age_yrs, sp) %>% summarise(n=n())
ggplot(d1) +
  geom_bar(aes(x=alz_age_yrs, y=n, fill=sp), stat="identity") +
  labs(title = "What age Alzerheim is diagnosed", subtitle = "by gender", caption = "") + 
  scale_fill_manual(values=c("dodgerblue3", "orchid2"))
```





## Medicines

```{r}
# most common medicines
# medicines bought 1-4 yrs before diagnose
# medicines bought 1 yr before diagnose

## how much reimbursements per years

## by year, what medicines people bought 
d1 <- drugs %>% 
  group_by(year(otpvm), substr(atc,1,3)) %>% 
  summarise(persons=length(unique(idnum)),
            plkm=sum(plkm),
            n=n())

d1$year <- d1$`year(otpvm)`
d1$atc3 <- d1$`substr(atc, 1, 3)`

# ggplot(d1)+
#   geom_line(aes(x=year, y=n, color=atc3))

ggplot(d1)+
  geom_line(aes(x=year, y=persons, color=atc3)) +
  labs(title="Number of persons who bought medicines (ATC3) by year ")

```


```{r}
d1 <- drugs %>% group_by(substr(atc, 1, 3)) %>% summarise(n=n())
d1$atc <- d1$`substr(atc, 1, 3)`
ggplot(d1) +
  geom_bar(aes(x=atc, y=n, fill=atc), stat="identity") +
  labs(title = "Most common medicine reimbursements (ATC3 level)")
```

```{r}
d1 <- drugs %>% 
  filter(substr(atc, 1, 3) == "N05") %>% 
  group_by(atc) %>% 
  summarise(n=n())

ggplot(d1) +
  geom_bar(aes(x=atc, y=n), stat="identity") +
  coord_flip() +
  labs(title = "ATC N05 medicine reimbursements")
```



## Care registry primary diagnoses

```{r}
# what other diagnoses people have 1yr before alz-diagnose
# 2-5yrs before alz-diganose

# Join person data to care register data
d <- left_join(persons, care_reg)
d1 <- d %>% filter(adm_date >= "1996-01-01")
# d1[d1$pdgo == "",] # is there empty, YES

d1$length_of_carereg <- d1$disch_date - d1$adm_date



```

### Within 1 year before diagnose

```{r}
# hoitojakso alkanut 1v sisällä, primary diagnose ICD-10
# lenght of care reg 
d2 <- d1 %>% 
  filter(erko307_pv - adm_date <= 365.25,  
         erko307_pv - adm_date >=0 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            persons=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
            )

d2_agg$mean_of_length <- as.integer(d2_agg$mean_of_length)

ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg, size=mean_of_length))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=mean_of_length, label = pdg, size=n ), 
            position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-persons, -n, -mean_of_length)
```




```{r}
## how many days in 
d2_agg <- d2 %>% 
  group_by(idnum) %>% 
  summarise(days=as.integer(sum(length_of_carereg)),
            n=n()
            )

ggplot(d2_agg) +
  geom_histogram(aes(days), bins = 20) +
  labs(title = "How many days in care?")
summary(d2_agg)
```





### Within 1-5 years before diagnose

```{r}


d2 <- d1 %>% 
  filter(erko307_pv - adm_date <= 365.25*5,  
         erko307_pv - adm_date >=365.25 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            persons=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
  )
d2_agg$mean_of_length <- as.integer(d2_agg$mean_of_length)


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg, size=mean_of_length))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=mean_of_length, label = pdg, size=n ), 
            position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-persons, -n, -mean_of_length)

```

### 5+ years before diagnose

- since 1996

```{r}

d2 <- d1 %>% 
  filter(erko307_pv - adm_date >=365.25*5 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            persons=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
  )
d2_agg$mean_of_length <- as.integer(d2_agg$mean_of_length)

ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=n, label = pdg, size=mean_of_length))


ggplot(d2_agg) +
  # geom_point(aes(x=persons, y=n)) +
  geom_text(aes(x=persons, y=mean_of_length, label = pdg, size=n ), 
            position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-persons, -n, -mean_of_length)

```



## Long-term institutionalizations

```{r}
length(unique(inst_reg$idnum))

d1 <- inst_reg
d1$kesto <- difftime(d1$lhjlakpv, d1$lhjalkpv, units = c("days"))

ggplot(d1) + 
  geom_histogram(aes(d1$kesto), bins = 20)
```



## Most common ICD10 primary and secondary diagnoses 

From care registry

```{r}
## ICD10 wordcloud 
# c(pdgo, pdge) # primary ICD-10
# c(sdg1o, sdg1e, sdg2o, sdg2o) # secondary ICD-10
# care_reg

care_reg %>% 
  select(idnum, adm_date, pdgo, pdge, sdg1o, sdg1e, sdg2o, sdg2o) %>% 
  pivot_longer(cols = c(pdgo, pdge, sdg1o, sdg1e, sdg2o, sdg2o), names_to = "Primary", values_to = "ICD10") -> d1

d1 <- d1 %>%
  filter(ICD10 != "") %>%  
  group_by(ICD10) %>% 
  summarise(n=n())

pal2 <- brewer.pal(8,"Dark2")
wordcloud(words = d1$ICD10, 
          freq = d1$n, 
          scale = c(8, .2),
          min.freq = 10,
          colors = pal2)
```




# Graphs


## Care register periods

```{r}

## Hoitojaksot -----

# Join person data to care register data
d <- left_join(persons, care_reg)
d1 <- d %>% filter(adm_date >= "1996-01-01")
# d1[d1$pdgo == "",] # is there empty, YES

d1$length_of_carereg <- d1$disch_date - d1$adm_date

## Mitä hoitojaksoja on henkilöllä 1v diagnoosia ennen? ----
# hoitojakso alkanut 1v sisällä, primary diagnose ICD-10
# lenght of care reg 

d2 <- d1 %>% 
  filter(erko307_pv - adm_date <= 365.25,  
         erko307_pv - adm_date >=0 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            ppl=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
            )

ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg, size= as.integer(mean_of_length)))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=as.integer(mean_of_length), label = pdg, size=n ), position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-ppl, -n, -as.integer(mean_of_length))


## Mitä hoitojaksoja henkilöllä on 1-5v ennen diagnoosia? -----

d2 <- d1 %>% 
  filter(erko307_pv - adm_date <= 365.25*5,  
         erko307_pv - adm_date >=365.25 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            ppl=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
  )

ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg, size= as.integer(mean_of_length)))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=as.integer(mean_of_length), label = pdg, size=n ), position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-ppl, -n, -as.integer(mean_of_length))


## Mitä hoitojaksoja henkiölllä on 5+ v ennen diagnoosia?

d2 <- d1 %>% 
  filter(erko307_pv - adm_date >=365.25*5 )
d2_agg <- d2 %>% 
  group_by(pdg) %>% 
  summarise(n=n(), 
            ppl=length(unique(idnum)),
            mean_of_length=mean(length_of_carereg)
  )

ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=n, label = pdg, size= as.integer(mean_of_length)))


ggplot(d2_agg) +
  # geom_point(aes(x=ppl, y=n)) +
  geom_text(aes(x=ppl, y=as.integer(mean_of_length), label = pdg, size=n ), position=position_jitter(width=1,height=1))


d2_agg %>% arrange(-ppl, -n, -as.integer(mean_of_length))

```


