---
title: "Alzheimer demo"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(Hmisc)
library(lubridate) 
library(wordcloud)
library(plotly)
library(survival)
library(survminer)

# Load datasets
load(file="./datasets/alzdatasets.RData", verbose=T)

# disable scientific number formatting
options(scipen=999) 
```




Patients
=================

Inputs {.sidebar}
--------------------------------------------------------

You can subgroup the population by age and gender (If population size goes under 10, subset will return the whole population).

```{r sidebar}
h3("Population")

sliderInput(inputId = "ageslider", 
            label = "Select age group:", 
            min = min(persons$alz_age_yrs), 
            max = max(persons$alz_age_yrs), 
            value = c(min(persons$alz_age_yrs), max(persons$alz_age_yrs)), 
            dragRange = TRUE)
radioButtons("gender", 
             label = "Gender:", 
             choices = c("All", "Male", "Female"), 
             selected = "All")


h3("Descriptives")

sliderInput(inputId = "min_freq", 
            label = "Diagnose frequency", 
            min = 20, 
            max = 5000, 
            value = 100)

selectInput(inputId = "atc_level", 
            label = "ATC level:", 
            choices = c("atc3", "atc4"), 
            selected = "atc3")

sliderInput(inputId = "medicines_gap", 
            label = "Gap between diagnose and medicine reimbursement", 
            min = -1000, 
            max = 1000, 
            value = c(-365,-1),
            dragRange = TRUE)

observeEvent(c(unique(event_data("plotly_click", source = "pop")$x)), {
  updateSliderInput(session, "ageslider", value = c(unique(event_data("plotly_click", source = "pop")$x), unique(event_data("plotly_click", source = "pop")$x)))
})

data_pop <- reactive({
  d1 <- persons %>% 
    filter(between(alz_age_yrs, input$ageslider[1], input$ageslider[2])) # filter by age
  # filter by gender
  if(input$gender != "All"){
    d1 <- d1 %>%
      filter(sp == input$gender)
  }
  # if n > 9 ok, else send whole pop
  if(nrow(d1) > 9) {
    return(d1)
  }else{
    return(persons)
  }
  
})

```

<!-- ### Selected -->


<!-- # ```{r} -->
<!-- # renderPrint({ -->
<!-- #   # plotly plot click event, pop -->
<!-- #   s <- unique(event_data("plotly_click", source = "pop")$x) -->
<!-- #   s -->
<!-- # }) -->
<!-- #  -->
<!-- # ``` -->


Row {.tabset}
-----------------------------------------------------------------------


### Patients

```{r populationGraphPlotly}

renderPlotly({ 
      d1 <- data_pop() %>% 
        group_by(alz_age_yrs, sp) %>% 
        summarise(n=n())
      p <- ggplot(d1) +
        geom_bar(aes(x=alz_age_yrs, y=n, fill=sp), stat="identity") +
        labs(title = paste0("Population (n=", nrow(data_pop()),")" ), 
             subtitle = "by gender", caption = "", x="") + 
        scale_fill_manual(values=c("dodgerblue3", "orchid2")) + theme(legend.position="none")
      ggplotly(p, source="pop")
    })
```


### Mortality

```{r}
renderPlotly({
  
  d1 <- data_pop() %>% 
    mutate(died_after_diag = kuolpv - erko307_pv,
           died_after_diag_cat = cut(as.integer(died_after_diag),
                                     breaks=c(-Inf, 180, 365, 730, 1095, Inf),
                                     labels = c("< half year", "0.5 - 1 year", "1-2 years", "2-3 years", "3+ years")
                                     )
           ) %>% 
    filter(!is.na(died_after_diag_cat)) %>% 
    group_by(died_after_diag_cat, sp) %>% 
    summarise(n=n())

  # mortality after diagnose
  ggplot(d1) +
    geom_bar(stat = "identity", aes(x=died_after_diag_cat, y=n, fill=sp)) + 
    coord_flip() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Mortality, days after the diagnose", x= "", y="") +
    scale_fill_manual(values=c("dodgerblue3", "orchid2"))
})
```

### Survival analysis (Kaplan-Meier)

```{r}

surv_object <- reactive({
  d <- data_pop()
  surv_object <- survival::Surv(time = d$survived, event = d$event)
  return(surv_object)
})
 
renderPlot({
  # predictors <- input$kaplanmeier
  fit1 <- survfit(surv_object() ~ 1, data = data_pop())
  # if(predictors == "sp") fit1 <- survfit(surv_object() ~ sp, data = data_pop())
  # summary(fit1)
  ggsurvplot(fit1, data = data_pop(), pval = TRUE)
})
```







Row 
-----------------------------------------------------------------------

### Descriptives: care registery diagnoses in whole timeline

```{r}
# Make the wordcloud drawing predictable during a session
wordcloud_rep <- repeatable(wordcloud)

renderPlot({ 
  d1 <- data_pop() %>%
    right_join(care_reg) %>% 
    filter(!is.na(alz_age)) %>% 
    mutate(pdgo = substr(pdgo, 1,3),
           pdge = substr(pdge, 1,3),
           sdg1o = substr(sdg1o, 1,3),
           sdg1e = substr(sdg1e, 1,3),
           sdg2o = substr(sdg2o, 1,3),
           sdg2o = substr(sdg2o, 1,3)) %>% ##### TODO THIS IS NEW, could be also in initial data
    select(idnum, adm_date, pdgo, pdge, sdg1o, sdg1e, sdg2o, sdg2o) %>% 
    pivot_longer(cols = c(pdgo, pdge, sdg1o, sdg1e, sdg2o, sdg2o), names_to = "Primary", values_to = "ICD10") %>% 
    filter(ICD10 != "") %>%  
    group_by(ICD10) %>% 
    summarise(n=n()) %>% 
    mutate(
      group = substr(ICD10, 1, 1)
    ) %>% 
    filter(n > input$min_freq)
  
  # palette
  basecolors <- colorRampPalette(brewer.pal(9, "Set1"))(20)
  colorlist <- basecolors[match(d1$group, unique(d1$group))]
  
  wordcloud_rep(words = d1$ICD10, 
                freq = d1$n, 
                scale = c(6, .2),
                colors = as.character(colorlist),
                ordered.colors=TRUE)
})

```

### Descriptives: medicines

```{r}
renderPlotly({
  
  ## Medicines atc percentages, before diagnose
  atc_level <- ifelse(input$atc_level == "atc3", 3, 4)
  
  drugs %>% 
    left_join(data_pop()) %>% 
    filter(!is.na(alz_age)) %>% 
    mutate(atc_code = substr(atc, 1, atc_level),
           gap = otpvm - erko307_pv) %>% 
    filter(gap >= input$medicines_gap[1], gap <= input$medicines_gap[2]) %>% 
    select(idnum, atc_code) %>% 
    group_by(atc_code) %>% 
    summarise(n=n(),
              uniq=length(unique(idnum)),
              time="before") %>% 
    mutate(
      percentage = 100 * uniq / nrow(data_pop())
    ) %>% 
    select(atc_code, time, percentage) -> d1
  
  ## add labels
  d1 <- d1 %>% left_join(atc_labels)
  
  ggplot(data = d1, aes(x=atc_code, y=percentage, fill=atc_code)) +
    geom_bar(stat="identity") +
    geom_text(aes(label = paste0(atc_code, " ", label), 
                  y = 50), 
              # position = position_dodge(0.9), 
              hjust = 0,
              vjust = 0) +
    coord_flip() + 
    scale_y_continuous(limits = c(0,100), breaks = c(0,25,50,75,100)) +
    labs(title=paste0("Medicine usage between ",input$medicines_gap[1], " and ", input$medicines_gap[2], " days before diagnose"), x= "", y="% of patients") + theme(legend.position="none")
  
  
})

```

Timeline
===================


Inputs {.sidebar}
--------------------------------------------------------

You can select from the timeline graph a bar according to certain year, which will show most common diagnoses on graph below. 

Scatter plot shows only first main diagnoses (pdgo).

```{r}
# h3("Timeline")
# 
# radioButtons(inputId = "tl_var",
#              label = "y-axis:",
#              choices = c("patients"),
#              selected = "patients"
#              )
# 
# h3("Diagnoses")
# 
# radioButtons(inputId = "tl_diagnose_x",
#              label = "x-axis:",
#              choices = c("patients", "patients_percentage", "hospitalization days"),
#              selected = "patients"
#              )
# 
# radioButtons(inputId = "tl_diagnose_y",
#              label = "y-axis:",
#              choices = c("patients", "patients_percentage", "hospitalization days"),
#              selected = "days"
#              )

```


```{r create_timeline_data}
tl <- reactive({
  tl <- timeline_init %>% 
    filter(idnum %in% data_pop()$idnum)
  tl_year <- NULL
  for(i in seq(from = -15, to = 15, by = 1) ){
    if(is.null(tl_year)){
      tl_year <- tl %>% 
        # filter( year_start <= i & year_end > i ) %>% # HUOM. tasaus tässä!
        filter( (year_end > i-1 & year_end <= i)| (year_start < i & year_start > i-1)  ) %>% # HUOM. tasaus tässä!
        group_by(idnum) %>% ##
        filter(source_int == max(source_int)) %>%  ###
        ungroup() %>% 
        group_by(source) %>% 
        summarise(.groups = 'drop',
                  time=i,
                  n=n(),
                  patients = length(unique(idnum)) #,
                  # ids=list(unique(idnum))
        )
    }else{
      tl_year <- tl_year %>% 
        rbind(
          tl %>% 
            # filter( year_start <= i & year_end > i ) %>% # HUOM. tasaus tässä!
            filter( (year_end > i-1 & year_end <= i)| (year_start < i & year_start > i-1)  ) %>% # HUOM. tasaus tässä!
            group_by(idnum) %>% ##
            filter(source_int == max(source_int)) %>%  ###
            ungroup() %>% 
            group_by(source) %>% 
            summarise(.groups = 'drop',
                      time=i,
                      n=n(),
                      patients = length(unique(idnum)) #,
                      # ids=list(unique(idnum))
            )
        )
    }
  }
  
  return(tl_year)
})

tl_diagnoses <- reactive({
  tl_diags <- NULL
  tl_data <- timeline_init %>% 
    filter(idnum %in% data_pop()$idnum)
  for(i in seq(-15, 15, by = 1)){
    if(is.null(tl_diags)){
      tl_diags <- tl_data %>% 
        filter(source_int == 3 & ( (year_end > i-1 & year_end <= i) | (year_start < i & year_start > i-1)  ) ) %>% # HUOM. rajaus klikin mukaan
        mutate(pdgo = substr(pdgo, 1,3)) %>% ##### TODO THIS IS NEW, could be also in initial data
        group_by(pdgo) %>% 
        summarise(
          .groups = 'drop',
          year = i,
          cases = n(),
          patients = length(unique(idnum)),
          days = sum(time_length(lhjlakpv+1 - lhjalkpv, "days"))
        ) %>% 
        mutate(
          patients_percentage = 100 * patients / nrow(data_pop())
        )  
    }else{
      tl_diags <- tl_diags %>% 
        rbind(
          tl_data %>% 
            filter(source_int == 3 & ( (year_end > i-1 & year_end <= i) | (year_start < i & year_start > i-1)  ) ) %>% # HUOM. rajaus klikin mukaan
            mutate(pdgo = substr(pdgo, 1,3)) %>% ##### TODO THIS IS NEW, could be also in initial data
            group_by(pdgo) %>% 
            summarise(
              .groups = 'drop',
              year = i,
              cases = n(),
              patients = length(unique(idnum)),
              days = sum(time_length(lhjlakpv+1 - lhjalkpv, "days"))
            ) %>% 
            mutate(
              patients_percentage = round(100 * patients / nrow(data_pop()), digits = 4)
            )
      )
    }
  }
  return(tl_diags)
})

```





Row {.tabset}
-----------------------------------------------------------------------

### Timeline

```{r plotly_timeline}
renderPlotly({
    p <- ggplot(data=tl()) +
      geom_bar(aes(x=time, 
                   y=patients, 
                   group=source, 
                   fill=source), 
               stat="identity") +
      labs(title = "Population timeline (Click the bar to get more detailed information of diagnoses)",
           x="Year according to diagnose date", 
           y="patients"
  )
  
  ggplotly(p, source="timeline_graph")
  
})
```


```{r}
# output$timelineTable <- DT::renderDataTable({
#   DT::datatable(tl(), rownames = FALSE)
#   })
# DT::dataTableOutput("timelineTable")
```





Row {.tabset}
-----------------------------------------------------------------------

### Scatter plot

```{r}
renderPlotly({
  ## Get click from timeline_graph
  s <- unique(event_data("plotly_click", source = "timeline_graph")$x)
  if(is.null(s)){
    s <- 0
    # title <- paste0("Most common diagnoses in whole timeline")
  }
  title <- paste0("Most common diagnoses in year ", s)
  
  ## xvar
  # if(input$tl_diagnose_x ==)

  ## Plot diagnoses scatter plot
  p <- ggplot(data=tl_diagnoses()[tl_diagnoses()$year %in% s, ]) +
    # geom_point(aes(x=patients, y=days, group=pdgo)) +
    geom_text(aes(x = patients,  
                  y = days, #get(input$tl_diagnose_y), 
                  group=pdgo, 
                  label = pdgo)) +
    labs(title = title, 
         x = "patients", 
         y = "hospitalization days" )#input$tl_diagnose_y)
  
  ## output
  p
})
```



### Table

```{r}
# renderPrint({
#  # plotly plot click event, pop
#  s <- unique(event_data("plotly_click", source = "timeline_graph")$x)
#  if(is.null(s)) s <- seq(-15,15,1)
#  s
# })

output$tl_dg_table <- DT::renderDataTable({
  ## Get click from timeline_graph
  s <- unique(event_data("plotly_click", source = "timeline_graph")$x)
  if(is.null(s)){
    s <- 0
    # title <- paste0("Most common diagnoses in whole timeline")
  }
  
  DT::datatable(tl_diagnoses()[tl_diagnoses()$year %in% s, ], 
                selection = list(
                  target = "row",
                  selected = c(1)),
                rownames = FALSE, 
                # colnames = c('Code range', 'Label'),
                options = 
                  list(pageLength = 5)
                )
  })
DT::dataTableOutput("tl_dg_table")

```



Diagnoses
==================


Row {.tabset}
-----------------------------------------------------------------------



**Diagnoses from care registry.** Cumulative number of cases. Same ICD-10 diagnose is counted only once by person. Diagnoses allocated to main (pdgo, pdge) and side diagnoses (sdg1o, sdg1e, sdg2o, sdg2o). You can select from the table below diagnoses to the graph.


```{r}
main_diagnoses <- reactive({
   # level out pdgo and pdge
  care_reg %>% 
    left_join(data_pop()) %>% 
    filter(!is.na(alz_age)) %>% 
    select(idnum, erko307_pv, adm_date, pdgo, pdge) %>%
    filter(pdgo != "") %>%
    pivot_longer(cols = c(pdgo, pdge), names_to = "diag_order", values_to = "ICD10") %>% 
    filter(ICD10 != "") %>% 
    mutate(date_mm = format(adm_date, "%Y-%m"),
           days_before =  adm_date - erko307_pv,
           icd=substr(ICD10, 1, 3) #this filters
    ) %>% 
    # filter( icd %in% unlist(icd10$classes[icd10$CodeRange %in% input$diagnoses])) %>%  #this filters
    group_by(idnum, ICD10) %>% 
    filter(adm_date == min(adm_date)) %>% 
    slice(1) %>% 
    ungroup() %>%
    arrange(ICD10, days_before) %>% 
    group_by(ICD10, days_before) %>% 
    summarise(n = n()) %>% 
    arrange(ICD10, days_before) %>% 
    mutate(
      cases = cumsum(n),
      diag_aggregated = substr(ICD10, 1, 3)
    ) %>% 
    rename(
      diag=ICD10
    ) -> d1
  
  return(d1)
})


side_diagnoses <- reactive({
   # level out pdgo and pdge
  care_reg %>%
    left_join(data_pop()) %>% 
    filter(!is.na(alz_age)) %>% 
    select(idnum, erko307_pv, adm_date, sdg1o, sdg1e, sdg2o, sdg2o) %>%
    filter(sdg1o != "") %>%
    pivot_longer(cols = c(sdg1o, sdg1e, sdg2o, sdg2o), names_to = "diag_order", values_to = "ICD10") %>%
    filter(ICD10 != "") %>%
    mutate(date_mm = format(adm_date, "%Y-%m"),
           days_before =  adm_date - erko307_pv,
          icd=substr(ICD10, 1, 3) #this filters
    ) %>% 
    # filter( icd %in% unlist(icd10$classes[icd10$CodeRange %in% input$diagnoses])) %>%  #this filters
    group_by(idnum, ICD10) %>%
    filter(adm_date == min(adm_date)) %>%
    slice(1) %>%
    ungroup() %>%
    arrange(ICD10, days_before) %>%
    group_by(ICD10, days_before) %>%
    summarise(n = n()) %>%
    arrange(ICD10, days_before) %>%
    mutate(
      cases = cumsum(n)
    ) %>%
    rename(
      diag=ICD10
    )
})

```



### Main diagnoses


```{r}
renderPlot({
  dplot <- main_diagnoses() 
  
  s <- input$diagnose_table_rows_selected
  classes <- as.character(unlist(icd10$classes[s]))
  if(!is.null(classes)){
    dplot <- dplot %>% 
      filter(substr(diag, 1, 3) %in% classes)
  }
  
  # Plot main diagnoses
  ggplot(dplot, aes(x = days_before, y = cases, group=diag)) +
    geom_line() +
    # scale_x_continuous() +
    geom_text(data = dplot %>% filter(days_before == last(days_before)), aes(label = diag, 
                                                                         x = days_before -300, 
                                                                         y = cases)) + 
    guides(color = FALSE) + 
    labs(title="ICD-10 main diagnoses") +
    geom_vline(xintercept=2, color="#667F66")
})
 


```


### Secondary diagnoses

```{r}

renderPlot({
  dplot <- side_diagnoses() 
  
  s <- input$diagnose_table_rows_selected
  classes <- as.character(unlist(icd10$classes[s]))
  if(!is.null(classes)){
    dplot <- dplot %>% 
      filter(substr(diag, 1, 3) %in% classes)
  }
  
  # Plot side diagnoses
  ggplot(dplot, aes(x = days_before, y = cases, group=diag)) +
    geom_line() +
    # scale_x_continuous() +
    geom_text(data = dplot %>% filter(days_before == last(days_before)), aes(label = diag,
                                                                         x = days_before -300,
                                                                         y = cases)) +
    guides(color = FALSE) +
    labs(title="ICD-10 secondary diagnoses") +
    geom_vline(xintercept=2, color="#667F66")
})

```


Row 
-----------------------------------------------------------------------

```{r}
output$diagnose_table <- DT::renderDataTable({
  DT::datatable(icd10[, c("CodeRange", "Description")] , 
                selection = list(
                  target = "row",
                  selected = c(1)),
                rownames = FALSE, 
                colnames = c('Code range', 'Label'),
                options = 
                  list(pageLength = 10)
                )
  })
DT::dataTableOutput("diagnose_table")
```



Medicines 
==================


Inputs {.sidebar}
--------------------------------------------------------

**Medicine reimbursements from Kela registry.** The Anatomical Therapeutic Chemical (ATC)  Classification System defines active ingredients of drugs to classes.  Timeline graph shows cumulative number of events when patients bought medicines. (Only reimbursement events are counted, not how many packages patience bought in the event.)

You can select one or multiple ATC classes from table below. You can also select hierarchy level for overall plot.

```{r}
radioButtons(inputId = "atc_barclass",
             label = "ATC level:",
             choices = c("ATC3", "ATC4", "ATC5", "ATC6", "ATC7"),
             selected = "ATC5"
             )
```


Row {.tabset}
-----------------------------------------------------------------------


### Timeline

```{r}
drugs_data <- reactive({
   drugs %>% 
    left_join(data_pop()) %>% 
    filter(!is.na(alz_age)) %>% 
    mutate(days_before = otpvm - erko307_pv,
           atc3=substr(atc,1,3)
    ) -> drugs_days_before
  
  drugs_days_before %>%
    arrange(atc, days_before) %>% 
    group_by(atc, days_before) %>% 
    summarise(n = n()) %>% 
    arrange(atc, days_before) %>% 
    mutate(
      cases = cumsum(n),
      atc3 = substr(atc, 1,3)
    )-> d1
  
  return(d1)
})


renderPlot({
  
  s <- input$atc_table_rows_selected
  codes <- as.character(unlist(atc_codes$atc[s]))
  
  plot_data <- drugs_data() 
  if(!is.null(codes)){
    plot_data <- plot_data %>% filter(atc3 %in% codes)
  }

  ggplot(plot_data, aes(x=days_before, y=cases, group=atc, color=atc3)) +
    geom_line() +
    # scale_x_continuous(trans = "reverse") +
    geom_text(data = plot_data %>% filter(days_before == last(days_before)),
              aes(label = atc, x = days_before -300, y = cases)) +
    guides(color = FALSE) +
    labs(title="cases") +
    geom_vline(xintercept=2, color="#667F66")
})

```

### Overall

```{r}
renderPlotly({
  
  ## Medicines atc percentages, before diagnose
  atc_level <- substr(input$atc_barclass, 4, 5)
    
  s <- input$atc_table_rows_selected
  codes <- as.character(unlist(atc_codes$atc[s]))

  
  drugs %>% 
    left_join(data_pop()) %>% 
    filter(!is.na(alz_age)) %>% 
    mutate(atc3=substr(atc,1,3),
           atc_code = substr(atc, 1, atc_level),
           gap = otpvm - erko307_pv) %>% 
    filter(atc3 %in% codes) %>% 
    filter(gap >= input$medicines_gap[1], gap <= input$medicines_gap[2]) %>% 
    select(idnum, atc_code) %>% 
    group_by(atc_code) %>% 
    summarise(n=n(),
              uniq=length(unique(idnum)),
              time="before") %>% 
    mutate(
      percentage = 100 * uniq / nrow(data_pop())
    ) %>% 
    select(atc_code, time, percentage) -> d1
  
  ## add labels
  d1 <- d1 %>% left_join(atc_labels)
  d1$label_name <- paste0(d1$atc_code, " ", d1$label)
  
  # palette
  # basecolors <- colorRampPalette(brewer.pal(7, "Accent"))(123)
  # colorlist <- basecolors[match(d1$atc3, unique(d1$atc3))]
  
  ggplot(data = d1, aes(x=label_name, y=percentage, fill=atc_code)) +
    geom_bar(stat="identity") +
    # geom_text(aes(label = paste0(atc_code, " ", label), 
    #               y = 50), 
    #           # position = position_dodge(0.9), 
    #           hjust = 0,
    #           vjust = 0) +
    coord_flip() + 
    # scale_y_continuous(limits = c(0,100), breaks = c(0,25,50,75,100)) +
    # scale_fill_manual(name = "label_name", values = my_colors) +
    labs(
      title=paste0(""),
      x="",
      y="percentage (%) of selected population") + 
    theme(legend.position="none")
  
  
})
```




Row
----------------------------------

```{r}
output$atc_table <- DT::renderDataTable({
  DT::datatable(atc_codes, 
                selection = list(
                  target = "row",
                  selected = c(1,2,3,4,5,6,7)),
                rownames = FALSE, 
                # colnames = c('Code range', 'Label'),
                options = 
                  list(pageLength = 7)
                )
  })
DT::dataTableOutput("atc_table")

```



About
===================

Purpose of this Shiny App is to demonstrate interactive usage of hospital datasets in explorative manner. All datasets are synthetic, but represent real world data. Found a bug? If you know github, you can open an [issue](https://github.com/pohjois-savon-tietoallas/demo-alzheimer-shiny/issues). Otherwise contact the [creators](mailto:jani.miettinen@uef.fi?subject=I found a bug in Alzdemo).
</br>
**Datasets** are synthetic, but represent real world data. Original datasets are not freely available. If you are interested of the datasets, please contact [Reijo Sund](mailto:reijo.sund@uef.fi?subject=About alzdemo datasets).
</br>
**Source code:** <https://github.com/pohjois-savon-tietoallas/demo-alzheimer-shiny>
</br>
**Licence:** [CC BY](https://creativecommons.org/licenses/by/4.0/). This license allows reusers to distribute, remix, adapt, and build upon the material in any medium or format, so long as attribution is given to the creator. The license allows for commercial use.
</br>
Copyright (c) 2020 Jani Miettinen </br>
</br>